---
title: "homework 1"
author: "Dominik Pegler"
date: '2022-04-02'
output: html_document
---

Due: April 20 2022

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import data and join individual text files

```{r}
library(ggplot2)

filelist = list.files(pattern = ".*.txt")
datalist = lapply(filelist, function(x)read.table(x, header=T)) 
df <- do.call("rbind", datalist)
```

## data exploration

### structure

```{r}
str(df)
```

```{r}
summary(df)
```

### distribution of values

```{r}
table(unlist(df$cueSize))
```

```{r}
table(unlist(df$targetValidity))
```

```{r}
table(unlist(df$sbj))
```

```{r}
table(unlist(df$gender))
```

```{r}
table(unlist(df$age))
```

```{r}
ggplot(df,aes(age)) + geom_histogram(binwidth=1)
```

```{r}
table(unlist(df$answerCorrect))
```

```{r}
summary(df$reactionTime)
```

```{r}
ggplot(df,aes(reactionTime)) + geom_histogram(binwidth=0.1)
```

### density plot: reactionTime (condition = gender)

```{r}
ggplot(df,aes(reactionTime, fill=gender)) + geom_density(alpha=0.2)
```

### scatterplot: reactionTime vs cueSize

```{r}
ggplot(df,aes(age,reactionTime)) + geom_point(alpha=0.13)
```

## Transform variables

Data structure

```{r}
str(df)
```

```{r}
df$sbj <- factor(df$sbj)
```

```{r}
df$answerCorrect <- factor(df$answerCorrect,levels=c(0,1,99),labels=c("incorrect","correct","timeout"))
```

New data structure

```{r}
str(df)
```

## Data cleaning

eliminate participant with most incorrect answers

```{r}
df_incorrect <- df[df$answerCorrect == "incorrect",]
```

subjects ordered by count of incorrect numbers

```{r}
# count observations
sort(summary(df_incorrect$sbj),decreasing=TRUE)
```

```{r}
df_cleaned <- df[df$sbj != "12",]
```

```{r}
nrow(df)
```

```{r}
nrow(df_cleaned)
```

## Bonus: Calculate error rate per subject

new column for rate of not correct answers (incorrect + time-out)

including all participants

```{r}
summary(df$sbj)
```

All subjects have responded 320 times

```{r}
not_correct <- df[df$answerCorrect != "correct",]

sort(summary(not_correct$sbj) / 320, decreasing = TRUE)
```

Subjects 4, 12 (previously excluded) and 9 have the 3 highest error rates with approx. 10%

## Data cleaning II

eliminate participants with
* too long reaction time (time-out)  
* reaction time < 0.150s  
* reaction time > 1.000s

```{r}
df_cleaned_2 <- df_cleaned[df_cleaned$answerCorrect!="timeout" & df_cleaned$reactionTime<=1 & df_cleaned$reactionTime >= 0.15,]

nrow(df_cleaned_2)
```

**Percentage** of observations excluded by this measure

```{r}
(nrow(df_cleaned) - nrow(df_cleaned_2)) / nrow(df_cleaned) * 100
```

**Percentage** of change in mean reaction time as a consequnce of this measure

```{r}
(mean(df_cleaned$reactionTime) - mean(df_cleaned_2$reactionTime)) / mean(df_cleaned$reactionTime) * 100
```

Mean in reaction time decreased by 0.82% as a consequence

## Split data set and compare means

Split into incorrect and correct answers

```{r}
df_cleaned_2_correct <- df_cleaned_2[df_cleaned_2$answerCorrect=="correct",]
nrow(df_cleaned_2_correct)
```

```{r}
df_cleaned_2_incorrect <- df_cleaned_2[df_cleaned_2$answerCorrect=="incorrect",]
nrow(df_cleaned_2_incorrect)
```

### Means

```{r}
aggregate(df_cleaned_2_correct$reactionTime, list(df_cleaned_2_correct$cueSize, df_cleaned_2_correct$targetValidity), FUN=mean) 
```

```{r}
aggregate(df_cleaned_2_incorrect$reactionTime, list(df_cleaned_2_incorrect$cueSize, df_cleaned_2_incorrect$targetValidity), FUN=mean) 
```

## Bonus: Validity effect

1) per subject for all conditions 
2) total for all conditions

```{r}
df_cleaned_2_correct
```


```{r}
df_cleaned_2_correct[, c(cueCondition, targetValidity, sbj)]
```


data_validity <- dcast(df_cleaned_2_correct_means_sbj, sbj + cueCondition ~ targetValidity, value.var = "meanRT")

head(data_validity) # zeigt die erste sechs Zeilen
```


```{r}
valid <- df_cleaned_2_correct[df_cleaned_2_correct$targetValidity=="valid",]
valid <- aggregate(valid$reactionTime, list(valid$sbj), FUN=mean) 
valid
```

```{r}
invalid <- df_cleaned_2_correct[df_cleaned_2_correct$targetValidity=="invalid",]
invalid <- aggregate(invalid$reactionTime, list(invalid$sbj), FUN=mean) 
invalid
```

Mean difference per subject

```{r}
invalid$x - valid$x
```

Overall mean difference

```{r}
mean(invalid$x - valid$x)
```

