---
title: "homework 3"
author: "Dominik Pegler"
date: '2022-06-05'
output: html_document
---

Due: June 05 2022

## Prepare data

```{r}
library(data.table)
library(ggplot2)

filelist = list.files(pattern = ".*.txt")
datalist = lapply(filelist, function(x)read.table(x, header=T)) 
df <- do.call("rbind", datalist)
setDT(df) # Convert data.frame to data.table

df$sbj <- factor(df$sbj)
df$answerCorrect <- factor(df$answerCorrect,levels=c(0,1,99),labels=c("incorrect","correct","timeout"))

# eliminate subject 12
df <- df[df$sbj != "12",]
df <- df[df$answerCorrect=="correct" & df$reactionTime<=1 & df$reactionTime >= 0.15,]
```

**Simulate the design and data of your experiment (factors and number of trials per experimental condition) 1000 times to estimate the statistical power under the following conditions.**

1. Determine the statistical power of your experiment to find a validity effect of 20 ms.
2. How small can the validity effect be to achieve a test power of 80%.
3. How small can the validity effect be to achieve a test power of 90%.
4. Finds the minimum number of participants needed to find an effect of 20 ms with a test power of 90%. 
5. Finds the average minimum number of valid trials per cue condition needed to find an effect of 20 ms with a test power of 90%. To do this, you can adjust the trialChangeFactor variable to change the number of trials per participant. The variable is a multiplier, so a trialChangeFactor of 1.5 means one and a half times as many trials per participant as in your experiment, 0.5 means half as many trials, etc. The ratio of valid and invalid trials remains the same. You can calculate the average number of valid runs per person by determining this in your data set and multiplying it by the trialChangeFactor or by determining it from the simulated data.

Create variables `RTnormalized` and `RTnormalized_real`

```{r}
df$condition <- paste(df$targetValidity, df$cueSize, sep = "_")

# Variablen initiieren
vpn <- NULL
i <- NULL
j <- NULL
df$RTnormalized <- 0 

for (i in 1:length(unique(df$sbj))) { # Loop über alle Personen
  vpn_temp <- unique(df$sbj)[i] 
  for (j in 1:length(unique(df$condition))) { # Loop über alle Bedingungen pro Person (deshalb genestet [Loop im Loop])
    condition_temp <- unique(df$condition)[j] 
    meanRT_temp <- mean(df[condition == condition_temp & sbj == vpn_temp]$reactionTime) # mittlere Reaktionszeit für diese Bedingung ermitteln
    df[condition == condition_temp & sbj == vpn_temp]$RTnormalized <- 
      df[condition == condition_temp & sbj == vpn_temp]$reactionTime - meanRT_temp # mittlere Reaktionszeit von den Reaktionszeiten abziehen
  }
}

# Um die Reaktionszeiten wieder realistischer bzwl. leichter interpretierbar zu machen, addieren wir den Mittelwert aller Reaktionzeiten. Die Form der Verteilung bleibt davon unbeeinflusst.
df$RTnormalized_real <- df$RTnormalized + mean(df$reactionTime)

# plot distributions
effect <- 0.02 # minimal relevant effect size to simulate
condition <- rep(c("real", "normalized", "effect"),
                 each = length(df$RTnormalized_real))
reactionTimes <- c(df$reactionTime,
                   df$RTnormalized_real,
                   df$RTnormalized_real - effect)

rt_distributions <- data.table(condition = condition, reactionTimes = reactionTimes)

mean_rt <- rt_distributions[, .(meanRT = mean(reactionTimes)), by = .(condition)]

ggplot(rt_distributions, aes(reactionTimes, color = condition)) +
  geom_density(size = 1.5) +
  scale_color_viridis_d(end = 0.85) +
  theme(panel.background = element_blank(),
        panel.grid = element_line(color = "gray"),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank())
```

**Create template datatable**

```{r}
# copy from the real data used for the analysis
dt_template <- copy(df[, c("cueSize", "targetValidity", "sbj")])

# remove non-existing levels
dt_template$sbj <- factor(dt_template$sbj, levels = (unique(dt_template$sbj)))

# change sbj to 1:end
levels(dt_template$sbj) <- c(1:length(unique(dt_template$sbj)))
```

**The simulation script**

```{r}
runSimulation <- function(dt_template, Nsim, effect, sampleSize, simSampleSize, trialChangeFactor){
    simList <- vector(mode = "list", length = Nsim)
    j <- NULL
    i <-  NULL
    for (j in 1:Nsim) {
      start.time <- Sys.time()
      simData <- copy(dt_template)
      if (simSampleSize < sampleSize) {
        randomSbj <- sample(1:length(unique(simData$sbj)), simSampleSize)
        simData <- simData[simData$sbj %in% randomSbj,]
        levels(simData$sbj)[levels(simData$sbj) %in% randomSbj] <- c(1:simSampleSize)
        } else if (simSampleSize > sampleSize) {
          sampleSizeIncrease <- simSampleSize - sampleSize
          if (sampleSizeIncrease <= sampleSize) {
            randomSbj <- sample(1:length(unique(simData$sbj)), sampleSizeIncrease)
            simDataAddition <- simData[simData$sbj %in% randomSbj,]
            simDataAddition$sbj <- factor(simDataAddition$sbj, levels = (unique(simDataAddition$sbj)))
            levels(simDataAddition$sbj) <- c((sampleSize+1):(sampleSize+sampleSizeIncrease))
            simData <- rbind(simData, simDataAddition)
          } else if (sampleSizeIncrease > sampleSize) {
            simRepetition <- floor(simSampleSize / sampleSize)
            simDataBase <- simData[rep(seq_len(nrow(simData)), times = simRepetition), ]
            simDataBase$repetition <- rep(1:simRepetition, each = nrow(simData))
            simDataBase$sbj_new <- as.numeric(simDataBase$sbj)
            simDataBase$sbj_new <- simDataBase$sbj_new + sampleSize * (simDataBase$repetition-1)
            simDataBase$sbj <- as.factor(simDataBase$sbj_new)
            simDataBase$repetition <- NULL
            simDataBase$sbj_new <- NULL
            randomSbj <- sample(1:length(unique(simData$sbj)), simSampleSize %% sampleSize)
            simDataAddition <- simData[simData$sbj %in% randomSbj,]
            simDataAddition$sbj <- factor(simDataAddition$sbj, levels = (unique(simDataAddition$sbj)))
            levels(simDataAddition$sbj) <-
            c((length(unique(simDataBase$sbj))+1):(sampleSize+sampleSizeIncrease))
            simData <- rbind(simDataBase, simDataAddition)
          }
        }

      if (trialChangeFactor < 1) {
        simData <- simData[sample(1:nrow(simData), round(nrow(simData) * trialChangeFactor))]
      } else if (trialChangeFactor > 1) {
        simData <- rbind(simData[rep(seq_len(nrow(simData)), times = floor(trialChangeFactor)), ],
                         simData[sample(1:nrow(simData), round(nrow(simData) * (trialChangeFactor-floor(trialChangeFactor))))])
      }

      for (i in 1:simSampleSize) {
        # simulates random interindividual variance
        individualRTvariance <-  rnorm(1, mean = 0, sd = 0.05)
        RT_var_sd <- 0.015
        RT_var1 <- rnorm(1, mean = 0, sd = RT_var_sd)
        RT_var2 <- rnorm(1, mean = 0, sd = RT_var_sd)
        RT_var3 <- rnorm(1, mean = 0, sd = RT_var_sd)
        RT_var4 <- rnorm(1, mean = 0, sd = RT_var_sd)

        # simulate reaction times per condition
        RT_1 <- sample(df$RTnormalized_real - effect + individualRTvariance,
                       nrow(simData[sbj == i & targetValidity == "valid" & cueSize == "small"]),
                       replace = TRUE) + RT_var1
        RT_2 <- sample(df$RTnormalized_real + individualRTvariance,
                       nrow(simData[sbj == i & targetValidity == "valid" & cueSize == "big"]),
                       replace = TRUE) + RT_var2
        RT_3 <- sample(df$RTnormalized_real + individualRTvariance,
                       nrow(simData[sbj == i & targetValidity == "invalid" & cueSize == "small"]),
                       replace = TRUE) + RT_var3
        RT_4 <- sample(df$RTnormalized_real + individualRTvariance,
                       nrow(simData[sbj == i & targetValidity == "invalid" & cueSize == "big"]),
                       replace = TRUE) + RT_var4

        # Zuordnung der simulierten Reaktionszeiten zu den Faktorkombinationen
        simData[sbj == i & targetValidity == "valid" & cueSize == "small", reactionTime := RT_1]
        simData[sbj == i & targetValidity == "valid" & cueSize == "big", reactionTime := RT_2]
        simData[sbj == i & targetValidity == "invalid" & cueSize == "small", reactionTime := RT_3]
        simData[sbj == i & targetValidity == "invalid" & cueSize == "big", reactionTime := RT_4]
      }

      simData[, simNo := j] # ergänzt neue Variable mit der Nummer der Simulation
      simList[[j]] <- copy(simData) # speichert das fertige simulierte Datenframe in die Liste mit den Simulationen
      end.time <- Sys.time() # End-Zeit der Simulation eines Datensatzes
      time.taken <- end.time - start.time # Dauer der Simulation eines Datensatzes
    }

    simData <- rbindlist(simList)

    # Zuletzt vergleichen wir noch die Standardabweichungen der Reaktionzeiten in den Bedingungen der realen und simulierten Daten
    data_trimmed_means_sbj <- df[, .(meanRT = mean(reactionTime)),
                                           by = .(sbj, cueSize, targetValidity)]

    data_trimmed_sd <- data_trimmed_means_sbj[, .(SD = sd(meanRT)), by = .(targetValidity, cueSize)]

    simData_trimmed_means_sbj <- simData[, .(meanRT = mean(reactionTime)),
                                         by = .(sbj, targetValidity, cueSize, simNo)]

    simData_trimmed_sd_sim <- simData_trimmed_means_sbj[, .(SD = sd(meanRT)),
                                                        by = .(targetValidity, cueSize, simNo)]

    simData_trimmed_sd <- simData_trimmed_sd_sim[, .(SD = mean(SD)),
                                                 by = .(targetValidity, cueSize)]

    # Berechnung des Validitätseffeks
    simData_means_sbj <- simData[, .(meanRT = mean(reactionTime)),
                                 by = .(simNo, targetValidity, cueSize, sbj)]

    simValidityEffect <- simData_means_sbj[, .(validityEffect = meanRT[targetValidity == "invalid"] -
                                                                meanRT[targetValidity == "valid"]),
                                           by = .(cueSize, sbj, simNo)]

    # Ergebnisse berechnen
    simValidityEffect_results <- simValidityEffect[,
                                     .(meanValidityEffect = t.test(validityEffect, mu = 0)$estimate,
                                       p.value = t.test(validityEffect, mu = 0)$p.value,
                                       t.stat = t.test(validityEffect, mu = 0)$statistic),
                                     by = .(simNo, cueSize)]

    sigNiveau <- 0.05 # Signifikanzniveau definieren

    powerResults <- simValidityEffect_results[t.stat > 0 & p.value < sigNiveau,
                                              .(EffectFound = .N,
                                                Nsim = max(simValidityEffect_results$simNo),
                                                testPower = .N / max(simValidityEffect_results$simNo)),
                                              by = .(cueSize)]

    return(powerResults)}
```

---

## 1. Determine the statistical power of your experiment to find a validity effect of 20 ms.

**Simulate reaction times**

```{r}
Nsim <- 1000 # Number of simulations
effect <- 0.02 # 20 ms validity effect (smallest relevant effect size)
sampleSize <- length(unique(dt_template$sbj)) # sample size of data
simSampleSize <- sampleSize # sample size for simulation (can be changed!)
trialChangeFactor <- 1 # change factor for trial number to increase or decrease trial number

runSimulation(dt_template, Nsim, effect, sampleSize, simSampleSize, trialChangeFactor)
```

Power for an effect o 20ms in the small cue size condition is 75% and 2.5% in the big cue size condition.

## 2. How small can the validity effect be to achieve a test power of 80%

```{r}
Nsim <- 1000 # Number of simulations
effect <- 0.0217 # 20 ms validity effect (smallest relevant effect size)
sampleSize <- length(unique(dt_template$sbj)) # sample size of data
simSampleSize <- sampleSize # sample size for simulation (can be changed!)
trialChangeFactor <- 1 # change factor for trial number to increase or decrease trial number

runSimulation(dt_template, Nsim, effect, sampleSize, simSampleSize, trialChangeFactor)
```

Around 21.7 ms in the condition with small cue size.

## 3. How small can the validity effect be to achieve a test power of 90%

```{r}
Nsim <- 1000 # Number of simulations
effect <- 0.025 # 20 ms validity effect (smallest relevant effect size)
sampleSize <- length(unique(dt_template$sbj)) # sample size of data
simSampleSize <- sampleSize # sample size for simulation (can be changed!)
trialChangeFactor <- 1 # change factor for trial number to increase or decrease trial number

runSimulation(dt_template, Nsim, effect, sampleSize, simSampleSize, trialChangeFactor)
```

Around 25 ms for the condition with small cue size.

## 4. Find the minimum number of participants needed to find an effect of 20 ms with a test power of 90%

```{r}
Nsim <- 1000 # Number of simulations
effect <- 0.02 # 20 ms validity effect (smallest relevant effect size)
#sampleSize <- length(unique(dt_template$sbj)) # sample size of data
simSampleSize <- 23 # sample size for simulation (can be changed!)
trialChangeFactor <- 1 # change factor for trial number to increase or decrease trial number

runSimulation(dt_template, Nsim, effect, sampleSize, simSampleSize, trialChangeFactor)
```

23 People for the condition with small cue size.

## 5. Find the average minimum number of valid trials per cue ...
... condition needed to find an effect of 20 ms with a test power of 90%. To do this, you can adjust the trialChangeFactor variable to change the number of trials per participant. The variable is a multiplier, so a trialChangeFactor of 1.5 means one and a half times as many trials per participant as in your experiment, 0.5 means half as many trials, etc. The ratio of valid and invalid trials remains the same. You can calculate the average number of valid runs per person by determining this in your data set and multiplying it by the trialChangeFactor or by determining it from the simulated data.


```{r}
avg_valid_runs_per_sbj <- mean(aggregate(df[df$targetValidity == "valid",sbj], list(df[df$targetValidity == "valid",sbj]), FUN = length)$x)
avg_valid_runs_per_sbj
```

```{r}
Nsim <- 1000 # Number of simulations
effect <- 0.02 # 20 ms validity effect (smallest relevant effect size)
sampleSize <- length(unique(dt_template$sbj)) # sample size of data
simSampleSize <- sampleSize # sample size for simulation (can be changed!)
trialChangeFactor <- 6 # change factor for trial number to increase or decrease trial number

runSimulation(dt_template, Nsim, effect, sampleSize, simSampleSize, trialChangeFactor)
```

```{r}
cat(avg_valid_runs_per_sbj * trialChangeFactor, "valid trials per person in the small cue size condition are necessary in order to reach 90% power to detect an effect of",1000*effect,"ms")
```

Simulation for the big cue size condition was not carried out since it would take extremely long given the still almost inexistent power for 6 times higher valid responses per subject. 

