---
title: "homework 4"
author: "Dominik Pegler"
date: '2022-06-12'
output: html_document
---

Due: June 12 2022

This home assignment is (partly) based on the previous ones. That is, only correct reaction times between 150 ms and 1 second are analyzed and the person with the most errors is excluded (data set "data_trimmed_correct").

In the final home assignment, we analyze the data to answer the research questions. These questions are, "Does the cue elicit a significant validity effect (significantly higher or lower than 0)? If so, in which direction is the validity effect (positive or negative validity effect)? Are the validity effects significantly different between the two cue conditions? One can analyze the validity effect directly. For example, this was done in task 3 of the 2nd home assignment. For this assignment, however, we analyze the trimmed correct reaction times. The following tasks are to be solved:

## Prepare data

```{r message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(lme4)
library(emmeans)
library(ggResidpanel)

filelist = list.files(pattern = ".*.txt")
datalist = lapply(filelist, function(x)read.table(x, header=T)) 
df <- do.call("rbind", datalist)
setDT(df) # Convert data.frame to data.table

df$sbj <- factor(df$sbj)
df$answerCorrect <- factor(df$answerCorrect,levels=c(0,1,99),labels=c("incorrect","correct","timeout"))

# eliminate subject 12
df <- df[df$sbj != "12",]
df <- df[df$answerCorrect=="correct" & df$reactionTime<=1 & df$reactionTime >= 0.15,]
```



## 1. Analyze ...
...the trimmed correct reaction times of your data set with mixed linear models (fixed factors: target validity and cue condition; random factor: random per-participant intercept). In doing so, start with the interaction model and gradually remove parameters. Find and describe the most parsimonious model that best fits the data.

Step 1: Remove interaction term
```{r}

rt.interaction <- lmer(reactionTime ~ cueSize * targetValidity + (1 | sbj),
                       df)

rt.main <- lmer(reactionTime ~ cueSize + targetValidity + (1 | sbj),
                df)

anova(rt.interaction, rt.main) # if no significant difference, we go with the main model
```

There is a significant difference, the interaction model describes the data better. No further removal steps needed. 


## 2. Repeat the analysis of Task 1, ...
... but this time starts with the null model and adds parameters incrementally. Finds and describes the most parsimonious and best-fitting model to the data.

```{r}
rt.null <- lmer(reactionTime ~ (1 | sbj),
                df)

# models with 1 main effect

rt.targetValidity <- lmer(reactionTime ~ targetValidity + (1 | sbj),
                          df)
rt.cueSize <- lmer(reactionTime ~ cueSize + (1 | sbj),
                        df)
```

Step 1: We start by adding targetValidity to the model
```{r}
anova(rt.null, rt.targetValidity) 
```

We can observe that adding targetValidity improves the model. Just to compare, what would happen if add cueSize instead first

```{r}
anova(rt.null, rt.cueSize) 
```

Same can be observed with cueSize. Both are a good choice to be added to the model.

Step 3: We decide to add cueSize and compare this new model to a model that includes both main factors.

```{r}
anova(rt.cueSize, rt.main) 
```

We observe a significant improvement of model fit and include targetValidity into the model.

Step 4: Now we again compare the main effects model with the interaction model and we will again observe a significantly better model fit. We therefor choose the interaction model. 

```{r}
anova(rt.main, rt.interaction) 
```

## 3. Perform post-hoc tests ...
... of the fitting model (the same model should be found in Task 1 and 2) using the package emmeans() to describe the effect found.

```{r message=FALSE, warning=FALSE}
emmeans(rt.interaction, pairwise ~ targetValidity | cueSize)
```

## 4. Interpret the found result ...
... in words. 

**"The model with both main effects and its interaction describes the data best and most parsimoniously. There are significantly faster response times in the valid condition compared to the invalid condition for both cue sizes."**


## 5. Generates diagnostic plots ...
... of the model found and briefly describes whether the model assumptions are met.

```{r}
resid_panel(rt.interaction, plots = "all")
```

**We can observe a linear association between predicted and real response times. We also observe that the residual variance is independent from our predicted values. Further, the assumpetion of normal-distributed residuals is met as we can see in the Q-Q-plot and the histogram.**


